Oriented Gradient Histogram applied to P300 Detection

### 2017-12-04 article.tex
Otras modificaciones

The P300~\citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and though it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\citep{Farwell1988} as shown in Fig.~\ref{fig:p300matrix}, where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{openvibep300matrix.png}
\caption{The P300 Matrix used in the OpenVibe Open Source software~\citep{Renard2010}. Rows are labeled from $1$ to $6$ and columns from $7$ to $12$. Rows/columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

@@

In order to complete the plot from the pixels, the Bresenham \citep{Bresenham1965,Ramele2016} algorithm is used to interpolate straight lines between each pair of these consecutive pixels.

@@

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the $16$ blocks and creating a histogram of gradients.  This technique is based on Lowe's SIFT~\citep{Lowe2004} method, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate the histogram, the interval $[0-360]$ of possible angles is divided in $8$ bins, each one at $45$ degrees.
Then, the angles $\theta$  are $ \theta = \{0, 45, 90, 135, 180, 225, 270, 315\} $. For each spacial bin $ i,j = \{1,2,3,4\} $, corresponding to the indexes of each block $B_{i,j}$,  the orientations are accumulated in a  $3$-dimensional histogram $h$ through the following equation: %In each step, the histogram is computed by multiplying the length of the gradient vector by a Wang trilinear interpolation between the gradient angle and the bin, in order to give importance to the module of the gradient.

@@

\noindent  where $ |J(\mathbf{x})| $ is the norm of the gradient vector found at each block of the patch using finite differences, $\angle J(\mathbf{x}) $ is the angle of the gradient vector,  $\theta$ is the angle bin, $\mathbf{x}$ is a pixel from  the $i,j$-block $B_{i,j}$ and $ w_\mathrm{ang}(\cdot) $  and $ w_{ij}(\cdot) $ are linear interpolation functions used by Lowe and Vedaldi et al. in~\citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each block when $s = 1$.  As the patch is of size $4 \times 4$ and  $8$ bin angles are considered, a descriptor of $128$ dimension is obtained. It can be observed that in each step, the histogram is computed by multiplying by $ |J(\mathbf{x})| $ , so the method considers both, the magnitude and the orientation of the gradient vector.

@@

The classification is carried out by using a discriminative semi-supervised classification method, based on the Naive Bayes Nearest Neighbor (NBNN) \citep{Boiman2008}.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There are one row and one column that matches the selected letter by the user.
\item For $ k= {1-6} $: Obtain the descriptor from the image of the signal plot according to the method described in Section~\ref{Feature}, $ d^r_1, \dots  d^r_6 $ for rows and  $ d^c_1, \dots  d^c_6 $ 
\item For $ k= {7-12} $: Compute

@@

\noindent This procedure is repeated for the 6 columns.

\begin{equation}
\hat{c} = \arg \min_u \sum_{q \in NN_T(d^c_u)}^{} \left\lVert q -  d^c_u \right\rVert ^2,\; u = 7,\dots,12
\label{eq:multiclassification}
\end{equation}

>>>>

The P300~\citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and though it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\citep{Farwell1988}. Fig.~\ref{fig:p300matrix} shows an example of a Speller Matrix used in the OpenVibe Open Source software~\citep{Renard2010}, where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

\begin{figure}[H]
\centering
\includegraphics[width=.49\linewidth]{openvibep300matrix.png}
\caption{Example of a Speller Matrix.  Rows and columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

@@

In order to complete the plot from the pixels, the Bresenham \citep{Bresenham1965,Ramele2016} algorithm is used to interpolate straight lines between each pair of  consecutive pixels.

@@

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the $16$ blocks and creating a histogram of gradients.  This technique is based on Lowe's SIFT~\citep{Lowe2004} method, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate the histogram, the interval $[0-360]$ of possible angles is divided in $8$ bins, each one at $45$ degrees.
 For each spacial bin $ i,j = \{1,2,3,4\} $, corresponding to the indexes of each block $B_{i,j}$,  the orientations are accumulated in a  $3$-dimensional histogram $h$ through the following equation: %In each step, the histogram is computed by multiplying the length of the gradient vector by a Wang trilinear interpolation between the gradient angle and the bin, in order to give importance to the module of the gradient.

@@

\noindent  where $ \theta \in \{0, 45, 90, 135, 180, 225, 270, 315\} $, $ i,j \in \{1,2,3,4\} $, $ |J(\mathbf{x})| $ is the norm of the gradient vector found at each block of the patch using finite differences, $\angle J(\mathbf{x}) $ is the angle of the gradient vector,  $\theta$ is the angle bin, $\mathbf{x}$ is a pixel from  the $i,j$-block $B_{i,j}$ and $ w_\mathrm{ang}(\cdot) $  and $ w_{ij}(\cdot) $ are linear interpolation functions used by Lowe and Vedaldi et al. in~\citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each block when $s = 1$.  As the patch is of size $4 \times 4$ and  $8$ bin angles are considered, a descriptor of $128$ dimension is obtained. It can be observed that in each step, the histogram is computed by multiplying by $ |J(\mathbf{x})| $ , so the method considers both, the magnitude and the orientation of the gradient vector.

@@

The following step is classify the descriptors $\{d_{kp}, kp =1,\dots, 12\}$ in order to determine the correct row and column of the chosen letter en the Speller Matrix.   The classification is carried out by using a discriminative semi-supervised classification method, based on the Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There are one row and one column that matches the selected letter by the user.
\item For $ k= {1-6} $: Obtain the descriptor from the image of the signal plot according to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_1, \dots,  d^c_6 $. 
\item Compute

@@

\noindent This procedure is repeated for the columns.

\begin{equation}
\hat{c} = \arg \min_u \sum_{q \in NN_T(d^c_u)}^{} \left\lVert q -  d^c_u \right\rVert ^2,\; u = 7,\dots,12
\label{eq:multiclassification}
\end{equation}
where $q$ is a posible descriptor, $NN_T(d)$ is the set of descriptors nearest to $d$ and $\hat{r}$ and$$
